---
title: "Activity 3: Predicting Future Occupancy for the Island Mouse"
author: "Conor McGowan, Nicole Angeli, Anna Tucker"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(tidyverse)
```

## The Island Mouse  
  
The the island mouse *Zapus islandsonious* is species of mouse found only on Castaway Island. These mice require sandy beaches for foraging for dune beetles and dense grasses for raising litters. They are sensitive to loud noises and temperature extremes, and the main sources of mortality are volcanic gases, exposure, predation, and collection by humans. Populations on beach sites are also susceptible to large storm events such as hurricanes.

## Single state occupancy projection

We'll start by reading in data on the current state of the system. This dataset contains 8 columns: 

column name | description
----------- | -----------
`pop`       | population name
`popID`     | two-letter id
`type`      | inland (1) or beach (2)
`abund`     | current estimated abundance
`occ`       | current single-state occupancy
`occ_multi` | current multi-state occupancy
`cert`      | level of certainty in current state
`noise`     | anthropogenic noise, 0 (none) to 5 (high)


```{r load data}
dat <- read.csv("island_mouse.csv")
head(dat)
```

We'll use these starting values to project the future occupancy of each stream for the next 50 years. In order to incorporate appropriate uncertainty into our projections, we need to run each projection for several realizations. The more the better, but here we'll just use 100 for demonstration purposes. We also create data structures to hold the outputs of the simulations. We can create a data frame that contains two empty columns that we'll fill in during the simulation. The column `prob` will hold the true underlying occupancy probability and `occ` will hold the realized presence or absence of the species.

```{r setup sims}
reps = 100           # Number of realizations
years = 50          # Number of years
nsites = max(dat$site) # Number of sites

sim1 = data.frame(site = rep(rep(1:nsites, each = years), reps),
                   year = rep(rep(1:years, nsites), reps),
                   rep = rep(1:reps, each = nsites*years),
                   prob = NA)

sim1 = full_join(sim1, dat[,1:3], by = "site")  # add current state
sim1$occ[sim1$year > 1] = NA #remove occupancy status for future years
head(sim1)
```

# Scenario 1 - constant conditions

First, we'll assume that the conditions at each stream will remain the same over the next 20 years. The occupancy probability in each year is dependent on the noise level at each site and the threat due to extreme storms. Beach sites (`type == 2`) are more susceptible to hurricanes than inland sites (`type == 1`). In the past, extreme storms have hit approximately once every 10 years, so we'll set the probability of a storm each year to 0.1.


```{r sim 1}

storm_prob = 0.1

for(i in 1:nrow(sim1)){
  
    # randomly draw whether a bad storm occurred that year
    storm = rbinom(1, 1, storm_prob)
    
    sim1$prob[i] = plogis(4 - 3*storm*(sim1$type[i]-1) - sim1$noise[i])
    sim1$occ[i] = rbinom(1, 1, sim1$prob[i])
}
```

We can visualize the occupancy of each site over time in each realization.

```{r plot occ 1}
sitelab = sapply(1:nsites, function(x) paste("Site", x))
    
sim1 %>%
    mutate(lab = factor(sitelab[site], levels = sitelab)) -> sim1

ggplot(sim1, aes(x = year, y = as.factor(rep))) +
    geom_tile(aes(fill = as.character(occ)), alpha = 0.7) +
    facet_wrap(~lab) +
    scale_fill_manual(values = c("darkgray", "dodgerblue3"),
                      labels = c(0, 1),
                      name = "presence") +
    xlab("Year") + 
    ylab("") +
    theme_bw() +
    theme(strip.background = element_rect(fill = "white"),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```

You can see here the effect of the stochastic simulation process on projected stream occupancy. For each site, all of these realizations were projected using the same conditions, and those conditions are constant across years. Even still, we can see variation in observed presence or absence of chub due to random chance.


We can also summarize the results of this projection by looking at the proportion of realizations in which the streams are still occupied at the end of the projections. 

```{r results sim1}
sim1 %>%
    group_by(site, year, type) %>%
    summarize(prop_occ = mean(occ)) %>%
    ungroup() %>%
    ggplot(aes(x = year, y = prop_occ, group = as.character(site),
           col = as.character(type))) +
    scale_color_manual(values = c("dodgerblue3", "springgreen4"),
                       labels = c("inland", "beach"),
                       name = "Site type") +
    geom_line(alpha = 0.7) +
    theme_bw() +
    ylab("Proportion of realizations occupied") +
    xlab("Year") +
    theme(legend.position = c(0.9,0.2)) +
    ylim(c(0, 1))


```

*How would you describe the results of this simulation? Is there evidence that stream occupancy will decline, improve, or remain stable if conditions remain the same?*

*Does site type seem to have a strong influence on future occupancy? If so, why? If not, what do you think is driving the observed variation in projected occupancy?*


# Scenario 2 - changing environmental conditions

In most projections, we want to account for changing environmental conditions. For this project, we elicited expert opinion on both the effect of the two main threats on island mosue persistence and the probability of those threats changing for each site over time. 

As above, we will start by creating the data frame to hold our outputs in a format that will make it easy to summarize and visualize them. Unlike above, however, we will only assign the current state of the noise level for each stream for year 1; for all subsequent years, we will allow those states to be drawn randomly for each year.

```{r sim 2}
sim2 = data.frame(site = rep(rep(1:nsites, each = years), reps),
                   year = rep(rep(1:years, nsites), reps),
                   rep = rep(1:reps, each = nsites*years),
                   prob = NA,
                  storm_prob = NA)

sim2 = full_join(sim2, dat[,1:4], by = c("site"))  # add current state

sim2$noise[sim2$year > 1] = NA         # remove noise status after year 1
sim2$occ[sim2$year > 1] = NA

head(sim2)
```

Following a similar process as above, we will loop through each scenario and save the projected occupancy of each site. However, now we will also be allowing the noise level and storm probability to vary.

*Anthropogenic noise*
These mice are very sensitive to anthropogenic noise. Our expert panel predicted that noise levels across the island would increase if the radio broadcast station was built, with sites closer to the station experiencing greater noise levels. We asked an expert panel to estimate the probability that noise levels at any given site will increase, and they decided by consensus that the probability of noise increase was 5% (0.05). In these simulations, we assume that sites can only become more noisy each year, not less, and that the maximum noise level is 5.

*Extreme storms*
In our previous simulations, we kept storm probability at 0.10, which reflected past trends in the frequency of extreme storms hitting Castaway Island. However, under climate change projections, the frequency and strength of storms is expected to increase. In this next set of simulations, we let the probability of an extreme storm event vary for each year. The probability of extreme storms in any given year ranges from the current probability (0.1) to an increased probability (0.2).


```{r sim 2 contd}

# Determine storm probability for each year
storm_prob = runif(nrow(sim2), 0.10, 0.20)

# Determine noise level for each site in each year
noise_prob = 0.05                    # Probability of increased noise
noise_lev = matrix(NA, nrow = nsites, ncol = years) # To hold values
noise_lev[,1] = dat$noise           # Initial noise levels
for(i in 1:nsites){
  for(t in 2:years){
    noise_lev[i,t] <- ifelse(runif(1,0,1) > noise_prob,
                             ifelse(noise_lev[i,t-1] < 5,
                                    noise_lev[i,t-1]+1,
                                    noise_lev[i,t-1]), noise_lev[i,t-1])
  }
}

# Project occupancy under new conditions
for(i in 1:nrow(sim2)){
    
    sim2$noise[i] = noise_lev[sim2$site[i], sim2$year[i]]
    sim2$storm[i] = rbinom(1, 1, storm_prob[i])
    
    sim2$prob[i] = plogis(4 - 3*sim2$storm[i]*(sim2$type[i]-1) -
                            sim2$noise[i])
    sim2$occ[i] = rbinom(1, 1, sim2$prob[i])
}
```

The projected occupancy of each site:

```{r plot occ 2}

sim2 %>%
    mutate(lab = factor(sitelab[site], levels = sitelab)) -> sim2

ggplot(sim2, aes(x = year, y = as.factor(rep))) +
    geom_tile(aes(fill = as.character(occ)), alpha = 0.7) +
    facet_wrap(~lab) +
    scale_fill_manual(values = c("darkgray", "dodgerblue3"),
                      labels = c(0, 1),
                      name = "presence") +
    xlab("Year") + ylab("Realization") +
    theme_bw() +
    theme(strip.background = element_rect(fill = "white"))
```


```{r plot covars 2}

sim2 %>%
    group_by(site, year, type) %>%
    summarize(prop_occ = mean(occ)) %>%
    ungroup() %>%
    ggplot(aes(x = year, y = prop_occ, group = as.character(site),
               col = as.character(type))) +
    scale_color_manual(values = c("dodgerblue3", "springgreen4"),
                       labels = c("inland", "beach"),
                       name = "Site type") +    
    geom_line(alpha = 0.7) +
    theme_bw() +
    ylab("Proportion of realizations occupied") +
    xlab("Year") +
    theme(legend.position = c(0.9, 0.8)) +
    ylim(c(0, 1))



```

*How would you interpret the output of this simulation? Do you expect site occupancy to increase, decrease, or remain stable? What does this mean for the island mouse population?* 

We can use these simulations to look at the effect of any combination of projected threats due to storms or noise on stream occupancy. Follow this link [link] to an interactive webpage that will allow you to input different scenarios. Use this application to answer the following questions:
* Question 1
* Question 2



Because the location of the station is not yet determined, we will incorporate this potential variation in noise through stochastic simulations. We have estimated that no matter where the station is built, at least 2 sites will be nearby. Those nearby sites will have a greater future noise level than others. 


Now, each replicate will represent a different scenario for both which sites are affected by noise and the probability of extreme storms. To incorporate appropriate stochasticity, we will run 100 realizations of each.
